Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-24 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Solve units issue with spread rate
It turns out that one of the metadata files stored with data had
incorrect information on units. Also, the description of some variables
was incorrect. A root issue is that the actual raw data is all missing,
the data as committed by Rita already includes calculations such as
spread rate and is missing the raw flaming start and end times, etc. I
must assume those simple calculations are correct, I am sure they are,
but it would be best to have the raw data with stopwatch times.

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires.
  Multiple studies have established the existence of non-additive effects in
  the flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture dynamics vary across species with higher specific
  leaf area being associated with higher maximum absorption and lower litter
  bulk denisty being associated with faster drying rates. Species that produce
  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
  species) have faster rates of litter desiccation. We examined two axes of
  flammability (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Furthermore, we found that litter mixtures exhibited
  consistent positive non additivity in flame spread rate and that fuel
  consumption and flame spread rate both showed increasing positive non
  additivity with time since wetting.

@@

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn \cite{Kreye+Varner+Etal-2018. In particular, fine dead fuels respond very readily to environmental changes \cite{Nelson 2001}.  Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}. Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}. Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season. Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Nelson 2001}.

Fire scientists have long categorized fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}. A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite[eg]{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001} that include at least temperature and relative humidity and most aim to describe equilibrium conditions although some include precipitation and solar radiation effects as well. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and therefore fire behavior can depend on rate of drying following a wetting event. If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting. Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires.

In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}. Similarly, strong species specific effects on litter flammability are recorded for the forests tof the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a strong role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.  Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown. We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density). Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.

Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior. In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer. Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}. These different components of fire behavior have differing ecological effects. Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species. Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.

We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. Previous work described the flammability of these species in monoculture and in mixture for dry litter \cite{Magalhaes+Schwilk-2012}. {TODO- more on oaks?]  Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. We compared leaf litter traits with the measured moisture absorption and desorption rates. Finally, we tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed). I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from A. magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m. The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.

@@

We produced dry-down curves to determine maximum water retainability and the rate of moisture content loss (dessication rate) for each species and for four distinct mixtures. These curves allowed us to  determine the rate at which a species loses moisture (approximated as
exponential decay), and determine maximum water content following saturation
and gravity draining. Samples from the eight species (six replicates per
species) and four mixtures (5 replicates per mixture type) were placed in
baskets (approximately 45 x 45 cm by 15 cm tall) constructed of 1 cm aluminum
screen. Samples were 450 g of litter (dry weight) per trial, and litter depth
was standardized to 10 cm by changing the horizontal dimensions of the basket.
Weight was measured by using a balance sensitive to 0.1 g (model XS16001L,
Mettler Toledo, Columbus, OH). Samples were immersed in plastic storage bins
filled with water for 24 hours to fully saturate them. After saturation, the
baskets were removed from the water, allowed to drain for 3 minutes, and
samples weighed to measure maximum moisture content. The samples were then
allowed to dry in a partially controlled environment kept at 21 C temperature
and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated)
placed over the samples. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100 C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than10\% moisture.

@@

With only eight species in this study, our power to test leaf trait effects on the observed moisture dynamics is low, so we decided a priori to test the effect of one leaf trait likely to influence moisture absoorption, specific leaf area, and one litter trait known to directly effect flammability and likely to influence drying rate, litter bulk density. Specific leaf area was taken from leaf fresh leaf area and leaf dry mass from X individuals per species and was collected as part of an ealier study \cite{Magalhaes+Schwilk-2012}.  Litter bulk density was obtained in this experiment for each basket by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timeing so as to attempt to burn samples over the full range of moisture
contents from 100\% to below 10\% moisture. To expand our data on the dry end
of this gradient, we supplemented these data with data from an earlier
experiment on well-dried litter \cite{Magalhaes+Schwilk-2012}. This
supplemental data added five replicates per species at an oven dried moisture
content of $<$ 5\%. We assigned these replicates a time since wetting value of
144 hours equivalent to the longest time since drying for the wetted samples.
Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm
tall, in which leaf litter was placed and the table gently shaken to allow
settling. This design allows fire to reach constant flame spread rates
\cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced along
the apparatus allowed visual determination of maximum flame height. Litter was
ignited with a propane torch; maximum adiabatic flame temperature of 1899 C,
http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and
allowed to burn until extinction. If a species at a specific moisture level
failed to ignite after three attempts, no further attempts were made. A timer
was used to record time to ignition and duration of flaming combustion. The
flammability trials followed the methodology detailed in de Magalhães and
Schwilk \cite{Magalhaes+Schwilk-2012}.

@@

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and mass loss (\%),
calculated as the difference between fuel mass before and after flame
extinction. Some trials resulted only in smoldering combustion; for those
trials, a time to ignition and mass loss were recorded, but flame spread rate
and flame height were recorded as zero. Our thermocouple system failed or data
was lost for two days of burning trials, therefore we have omitted those data
from this analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated
that flame spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability.

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimizes wind and helps
regulate temperature and relative humidity. Temperature and relative humidity
were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman,
Boothwyn, PA). To minimize temperature and relative humidity variation, all
trials were conducted on clear days between 10 am and 3 pm.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these c oefficents we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.

We tested if one leaf trait (specific leaf area) and one litter trait (litter
bulk density) influenced moisture dynamics (maximum water retainability and
desiccation rate) by fitting linear models with species means as observations.
For each response variable we fit a single linear model with two predictors and
an interaction term.

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. Due to similar leaf traits, species within a genus tended to behave
similarly and the two scale-leaved genera in the Cupressaceae,
\emph{Calocedrus} and \emph{Sequoiadendron} also behaved similarly to one
another. Therefore, we used taxonomic groups (\emph{Abies}, Cupressaceae,
\emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and
treated these as fixed effects in our models. The effect of fuel moisture and
time since wetting on flammability parameters were modeled using mixed effect
models with fuel moisture or time since wetting and taxon as fixed effects and
average vapor pressure deficit in the burning room during that trial as a
nuisance covariate.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures we first grouped the
species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four broad groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean content of the three consituent species at that time since wetting. We
predicted flammability parameters based on the models fit to single species
data described above and then calculated an expected moisture content and
spread rate as the average of these three values for the mixture. The
expectation from the null model was that the difference between observed and
predicted values of flammability will be zero. We tested if there were
significant departures from the null (zero) with a mixed effects linear model.
For flammability parameters, we fit linear mixed effect models with the
observed-predicted flammability value as the dependent variable and used Wald
tests to test if the centered intercept was different than zero (overall non
additivity) and if the slope with time since wetting was different from zero
(non additivity changing with drying).

@@

There was some evidence for non additive in drying rate a tendency for litter mixtures. Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig. \ref{fig:mixture_obs_pred_mc} less than 1, p < 0.0001).

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack a mechanistic
  understanding of moisture synergies in mixed litter and of moisture effects
  on the flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures, to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Finally, we found that litter mixtures exhibited
  consistent non additivity in flame spread rate in fuel consumption. This non
  additivity became more positive with fuel drying. Litter mixtures burn with
  fire behavior more similar to that of the most flammable constituent species
  and this effect increases as fuels dry.}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \cite{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to describe equilibrium conditions and at least  include temperature and relative humidity although some include precipitation and solar radiation effects as well. Explicit models of fire behavior as a function of time since wetting could improve fire prediction, however. In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting. Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be driven by by leaf traits, ermegent litter litter traits, or both. Species specific effects on litter flammability are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a role in flammability feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fueled by plant litter, overstory tree composition drives variation in local fire behavior \cite{Schwilk+Caprio-2011} as a result of differential litter flammability cross tree species and as a result of synergistic interactions in multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability feedbacks may help explain historical fire frequency patterns, but current estimates of litter flammability in mixed conifer forests are based largely on burning dry litter.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter [CITE]. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}. Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.

Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior. In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer. Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}. These different components of fire behavior have differing ecological effects. Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species. Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. There is some evidence for non addiitvity of flammability in moist fuels and one study has found increasing non additivity with moisture \cite{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree species in a temperate fire-prone forest in California. We determined the maximum water retention and rate of moisture loss and examined how these vary in relation to two leaf and litter bed traits.  We then examined the influence of moisture on the flammability across these species and how flame spread rate and fuel consumption varied with time since wetting and by taxon. Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed. Previous work described the flammability of these species in monocultures and in mixtures for oven-dry litter \cite{Magalhaes+Schwilk-2012} and found consistent positive non additivity of flammability where fire behavior of mixtures was driven by the most flammable species. Here, we examined if this positive non additivity was maintained in mixtures across varying levels of fuel moisture. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects as seen in oven-dried litter. We tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. We chose eight dominant tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, we consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed). Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32\,km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from \emph{A. magnifica} which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10\,m. The litter was collect about 2\,m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk. All litter was air dried down to < 5\% fuel moisture for weighing and dividing into moisture trials. The moisture manipulations and subsequent burning trials were conducted in April-August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets (approximately 45 x 45\,cm by 15\,cm
tall) constructed of 1 cm aluminum screen. Samples were 450 g of litter (dry
weight) per trial, and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were
immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100\,C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than 10\% moisture.

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absoorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
species and was collected as part of a previous study
\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timeing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite after 30 seconds, no further attempt was made and if three samples of a species at a particular mositure level failed to ignite, no further attempts were made on that species/moisture combination and the replicate was allowed to continue drying. A timer was used to record total time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed in de
Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

@@

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and mass loss were recorded, but flame spread rate and flame height
were recorded as zero. Our thermocouple system failed or data was lost for two
days of burning trials, therefore we have omitted those data from this
analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability
to avoid examining multiple redundant dependent variables.

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and helped
regulate temperature and relative humidity. Temperature and relative humidity
were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman,
Boothwyn, PA). All trials were conducted on clear days between 10 am and 3 pm.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these c oefficents we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that species within a genus tended to behave similarly and the two scale-leaved genera in the Cupressaceae,
\emph{Calocedrus} and \emph{Sequoiadendron}, also behaved similarly to one
another. Therefore, to preserve power and focus on analyses we used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models. The effect of fuel moisture and
time since wetting on flammability parameters were modeled using linear
models with fuel moisture or time since wetting and taxon as fixed effects and
average vapor pressure deficit in the burning room during that trial as a
nuisance covariate. We used linear mixed models to predict spread rate, but
fuel consumption tends to exhibit threshold effects with few intermediate
values. Therefore, we used mixed generalized linear models with a beta
regression link with the `glmmADMB` package \cite{Skaug_Fournier_etal-2016} to
model percent fuel consumed as a function moisture content and of time since
wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped the
species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four broad groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean content of the three consituent species at that time since wetting. We
predicted flammability parameters based on the models fit to single species
data described above and then calculated an expected moisture content and
spread rate as the average of these three values for the mixture. The
expectation from the null model was that the difference between observed and
predicted values of flammability will be zero. We tested if there were
significant departures from the null (zero) with a mixed effects linear model.
For flammability parameters, we fit linear mixed effect models with the
observed-predicted flammability value as the dependent variable and used Wald
tests to test if the centered intercept was different than zero (overall non
additivity) and if the slope with time since wetting was different from zero
(non additivity changing with drying).

@@

There was evidence for non additive effects in drying rate. Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig. \ref{fig:mixture_obs_pred_mc} less than 1, p < 0.0001).
